### 2024-11-29 parts/starfree.tex
big corrections

\AP Given a number $\omega \in \Nat$, let us write
$\intro*\ModuloTypes[\omega]^k$ for the collection of pairs $(S, \vec{r})$
where $S \subseteq \set{1, \dots, k}$ and $r \in \set{0, \dots, \omega - 1}^k$.
To a tuple $\vec{x} \in \Nat^k$, one can associate its \intro{$\omega$-type}
$\intro*\moduloType[\omega](\vec{x})$ which is the pair $(S, \vec{r})$ where $S
= \set{i \in \set{1, \dots, k} \mid x_i \geq \omega}$ and $\vec{r} = (x_i \mod
\omega)_{i \in \set{1, \dots, k}}$.

\begin{lemma}
    \label{decompose-polynomial:lem}
    Let $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
    \kl{$\Rel$-polyregular function},
    where we fix the alphabet $\Sigma = \set{a_1, \dots, a_k}$.
    There exists a computable
    $\omega \in \Nat_{\geq 1}$,
    and computable 
    \kl{integer binomial polynomials} 
    $P_{(S,\vec{r})} \in \Rel[\seqof{X_i}{i \in S}]$ for $(S,\vec{r}) \in \ModuloTypes[\omega]^k$,
    such that for all $\vec{x} \in \Nat^k$,
    \begin{equation*}
        f\left(
            \prod_{i = 1}^k a_i^{x_i}
        \right)
        = P_{(S, \vec{r})}
        \left(
            \seqof{\floor{x_i / \omega}}{i \in S}
        \right)
        \text{ where } (S, \vec{r}) = \moduloType[\omega](\vec{x})
        \quad .
    \end{equation*}
\end{lemma}
\begin{proof}
    \textbf{TODO}. 
\end{proof}

\begin{theorem}
    \label{decidable-n-poly:thm}
    Let $k,d \in \Nat$, and $f \in \ZPoly[d]$ be \kl{commutative}
    over an alphabet of size $k$.
    Then, the following are equivalent:
    \begin{enumerate}
        \item \label{f-combinatorial:item} $f$ is \kl{$(k,\Nat)$-combinatorial},
        \item \label{f-npoly-combi:item} $f \in \NPoly[d]$,
    \end{enumerate}
    Furthermore, the properties are decidable,
    and conversions effective.
\end{theorem}
\begin{proof}
    The implication
    \cref{f-npoly-combi:item}
    $\implies$ \cref{f-combinatorial:item} immediately follows
    from 
    \cref{n-poly-combinatorics:lem} and \cref{decide-rat-poly-npoly:cor}.
    For the converse implication, we use \cref{decompose-polynomial:lem} 
    effectively compute an $\omega \in \Nat$ and \kl{integer binomial polynomials}
    $\seqof{P_{t}}{t \in \ModuloTypes[\omega]^k}$.

    Now, consider $(S, \vec{r}) \in \ModuloTypes[\omega]^k$. 
    Notice that for all vectors $\vec{x} \in (\Nat_{\geq 1})^k$, 
    the vector $(x_1 \omega \ind{S}(1) + r_1, \dots, x_k \omega \ind{S}(k) + r_k)$ has
    \kl{$\omega$-type} $(S, \vec{r})$.
    In particular, the following
    equality holds:
    \begin{equation*}
        f\left(\prod_{i = 1}^k 
            a_i^{ x_i \omega \ind{S}(i) + r_i}
        \right)
        =
        P_{(S, \vec{r})}(\seqof{x_i}{i \in S})
        \quad  \forall \vec{x} \in (\Nat_{\geq 1})^k
        \quad .
    \end{equation*}

    Let us therefore consider the \kl{pumping pattern} $q \colon \Nat^k \to
    \Sigma^*$ that is simply defined as $q(X_1, \dots, X_k) \defined \prod_{i =
    1}^k a_i^{ X_i \omega \ind{S}(i) + r_i}$. 
    Because $f$ is \kl{$(k,\Nat)$-combinatorial}, there exists an $\omega_2 \in
    \Nat$ and a \kl{strongly natural binomial polynomial} $P \in \Rel[X_1,
    \dots, X_k]$ such that $f \circ q = P$ over $(\omega_2 \Nat_{\geq 1})^k$.
    In particular, the following
    equality holds:
    \begin{equation*}
        P_{(S, \vec{r})}(\seqof{\omega_2 x_i}{i \in S})
        = 
        P(\omega_2 x_1, \dots, \omega_2 x_k)
        \quad  \forall \vec{x} \in (\Nat_{\geq 1})^k
        \quad .
    \end{equation*}
    Because $\omega_2 \neq 0$, we conclude that $P_{(S, \vec{r})} = P$
    and that both are polynomials over $\Rel[\seqof{X_i}{i \in S}]$,
    and furthermore that $P_{(S, \vec{r})}$
    is a \kl{strongly natural binomial polynomial}.

    To conclude that $f \in \NPoly$, one can compute $f$ by first computing
    the \kl{$\omega$-type} of the input which is possible using a
    \kl{polyregular function}, and then computing
    the suitable \kl{strongly natural binomial polynomials} which is possible
    in $\NPoly$ thanks to \cref{decide-rat-poly-npoly:cor}. The
    resulting composition belongs to $\NPoly$ thanks to
    \cref{stability-polyregular:lemma},
    and we conclude that $f \in \NPoly[d]$ because
    it has \kl{growth rate} at most $d$
    (\cref{polyregular-polynomial-growth:lemma}).

    Let us remark that we actually proved that the function $f \in \ZPoly[d]$
    belongs to $\NPoly[d]$
    if and only if its corresponding polynomials
    $P_{(S, \vec{r})}$ are \kl{strongly natural binomial polynomials} for all
    $(S, \vec{r}) \in \ModuloTypes[\omega]^k$. Because 
    \cref{decompose-polynomial:lem} is effective, this $\omega \in \Nat$ and polynomials
    $P_{(S, \vec{r})}$ can be computed from $f$, 
    and one can decide 
    whether a polynomial is \kl{strongly natural binomial polynomial} using
    \cref{decide-rat-poly-npoly:cor}. Hence, one can decide whether
    $f \in \NPoly[d]$.
\end{proof}

@@

    Let us now define the function $g \colon \Sigma^* \to \Rel$ as follows:
    \begin{equation*}
        g \colon w \mapsto \begin{cases}
            g_{a^n} (w) & \text{ if } \card[a]{w} = n \text{ for some } a \in \Sigma \text{ and } n \leq N_0 \\
            \translate{N_0}(P)(\seqof{\card[a]{w} - N_0}{a \in \Sigma}) & \text{ otherwise }
        \end{cases}
    \end{equation*}
    We conclude that $g$ is a \kl{star-free $\Nat$-polyregular function}, and by definition
    $f = g$.
\end{proof}

>>>>

\AP Let us now introduce a decomposition of \kl{commutative}
\kl{$\Rel$-polyregular functions} into \kl{integer binomial polynomials}. Given
a number $\omega \in \Nat$, let us write $\intro*\ModuloTypes[\omega]^k$ for
the collection of pairs $(S, \vec{r})$ where $S \subseteq \set{1, \dots, k}$
and $r \in \set{0, \dots, \omega - 1}^k$. To a tuple $\vec{x} \in \Nat^k$, one
can associate its \intro{$\omega$-type}, written
$\intro*\moduloType[\omega](\vec{x})$, which is the pair $(S, \vec{r})$ where
$S = \set{i \in \set{1, \dots, k} \mid x_i \geq \omega}$ and $\vec{r} = (x_i
\mod \omega)_{i \in \set{1, \dots, k}}$.

\begin{lemma}[restate=decompose-polynomial:lem,label=decompose-polynomial:lem]
    \proofref{decompose-polynomial:lem}
    Let $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
    \kl{$\Rel$-polyregular function},
    where we fix the alphabet $\Sigma = \set{a_1, \dots, a_k}$.
    There exists a computable
    $\omega \in \Nat_{\geq 1}$,
    and computable 
    \kl{integer binomial polynomials} 
    $P_{(S,\vec{r})} \in \Rat[\seqof{X_i}{i \in S}]$ for $(S,\vec{r}) \in \ModuloTypes[\omega]^k$,
    such that for all $\vec{x} \in \Nat^k$,
    \begin{equation*}
        f\left(
            \prod_{i = 1}^k a_i^{x_i}
        \right)
        = P_{(S, \vec{r})}
        \left(
            \seqof{\floor{x_i / \omega}}{i \in S}
        \right)
        \text{ where } (S, \vec{r}) = \moduloType[\omega](\vec{x})
        \quad .
    \end{equation*}
\end{lemma}

\begin{theorem}
    \label{decidable-n-poly:thm}
    Let $k,d \in \Nat$, and $f \in \ZPoly[d]$ be \kl{commutative}
    over an alphabet of size $k$.
    Then, the following are equivalent:
    \begin{enumerate}
        \item \label{f-combinatorial:item} $f$ is \kl{$(k,\Nat)$-combinatorial},
        \item \label{f-npoly-combi:item} $f \in \NPoly[d]$,
    \end{enumerate}
    Furthermore, the properties are decidable,
    and conversions effective.
\end{theorem}
\begin{proof}
    Let $f \in \ZPoly[d]$ be \kl{commutative} over an alphabet of size $k$.
    We apply \cref{decompose-polynomial:lem} to compute an $\omega \in \Nat$ and
    \kl{integer binomial polynomials} $\seqof{P_{(S,\vec{r})}}{(S,\vec{r}) \in
    \ModuloTypes[\omega]^k}$ such that for all $\vec{x} \in \Nat^k$,
    $f\left(\prod_{i = 1}^k a_i^{x_i}\right) = P_{(S, \vec{r})}(\seqof{\floor{x_i / \omega}}{i \in S})$,
    where $(S, \vec{r}) = \moduloType[\omega](\vec{x})$.
    We are first going to prove that $f \in \NPoly[d]$ if and only if $P_{(S,
    \vec{r})}$ is a \kl{strongly natural binomial polynomial} for all $(S,
    \vec{r}) \in \ModuloTypes[\omega]^k$. This will also provide decidability
    of \cref{f-npoly-combi:item}, since one can decide whether a
    polynomial is \kl{strongly natural binomial polynomial} using
    \cref{decide-rat-poly-npoly:cor}.

    Assume that $f \in \NPoly$, then by definition, the polynomials $P_{(S,
    \vec{r})}$ are \kl{represented} by an \kl{$\Nat$-polyregular function},
    hence are \kl{strongly natural binomial polynomials}
    (\cref{decide-rat-poly-npoly:cor}). Conversely, if
    $P_{(S, \vec{r})}$ is a \kl{strongly natural binomial polynomial} for all
    $(S, \vec{r}) \in \ModuloTypes[\omega]^k$, then $f \in \NPoly$ because one
    can compute the \kl{$\omega$-type} of the input using a \kl{polyregular
    function}, and then compute the suitable \kl{strongly natural binomial
    polynomial} $P_{(S, \vec{r})}$ which is possible in $\NPoly$ thanks to
    \cref{decide-rat-poly-npoly:cor}. The resulting composition
    belongs to $\NPoly$ thanks to
    \cref{stability-polyregular:lemma},
    and we conclude that $f \in \NPoly[d]$ because it has \kl{growth rate} at
    most $d$
    (\cref{polyregular-polynomial-growth:lemma}).

    Note that the same proof scheme can be used to conclude that
    \cref{f-npoly-combi:item} implies
    \cref{f-combinatorial:item}. For the converse implication, we are
    going to introduce $\omega_2$ associated to the fact that $f$ is
    \kl{$(k,\Nat)$-combinatorial}. Because polynomials \kl{represented} by
    \kl{$\Nat$-polyregular functions} and \kl{integer binomial polynomials} are
    both closed under multiplication by a constant factor, we can assume that
    $\omega = \omega_2$ in the decomposition of $f$.
    Now, consider $(S, \vec{r}) \in \ModuloTypes[\omega]^k$. 
    Notice that for all vectors $\vec{x} \in (\Nat_{\geq 1})^k$, 
    the vector $(x_1 \omega \ind{S}(1) + r_1, \dots, x_k \omega \ind{S}(k) + r_k)$ has
    \kl{$\omega$-type} $(S, \vec{r})$.
    In particular, the following
    equality holds:
    \begin{equation*}
        f\left(\prod_{i = 1}^k 
            a_i^{ x_i \omega \ind{S}(i) + r_i}
        \right)
        =
        P_{(S, \vec{r})}(\seqof{x_i}{i \in S})
        \quad  \forall \vec{x} \in (\Nat_{\geq 1})^k
        \quad .
    \end{equation*}

    Let us therefore consider the \kl{pumping pattern} $q \colon \Nat^k \to
    \Sigma^*$ that is simply defined as $q(X_1, \dots, X_k) \defined \prod_{i =
    1}^k a_i^{ X_i \ind{S}(i) + r_i}$. 
    Because $f$ is \kl{$(k,\Nat)$-combinatorial} with parameter $\omega$,
    there exists a \kl{strongly natural binomial polynomial} $P \in \Rat[X_1,
    \dots, X_k]$ such that $f \circ q (\omega X_1, \dots, \omega X_k) = P (X_1, \dots, X_k)$ 
    over $(\Nat_{\geq 1})^k$.
    This proves that
    $P_{(S, \vec{r})}(\seqof{X_i}{i \in S})$
    equals
    $P(X_1, \dots, X_k)$ as polynomials,  hence,
    that $P_{(S, \vec{r})}$ is a \kl{strongly natural binomial polynomial}
    for all $(S, \vec{r}) \in \ModuloTypes[\omega]^k$. We have proven that
    $f \in \NPoly[d]$.
\end{proof}

@@

    Let us conclude by computing $f$ 
    using the following \kl{star-free $\Nat$-polyregular function} $g \colon \Sigma^* \to \Rel$:
    \begin{equation*}
        g \colon w \mapsto \begin{cases}
            g_{a^n} (w) & \text{ if } \card[a]{w} = n \text{ for some } a \in \Sigma \text{ and } n \leq N_0 \\
            \translate{N_0}(P)(\seqof{\card[a]{w} - N_0}{a \in \Sigma}) & \text{ otherwise }
        \end{cases}
        \qedhere
    \end{equation*}
\end{proof}
